---
title: "ADAGE signature analysis - case study"
author: "Jie Tan"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{User input example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This script shows how we analyze the *∆anr* mutation dataset used as a case study
in the ADAGE signature analysis paper. 

# Data preparation
Load in required libraries.

```{r}
library("ADAGEpath")
library("DT")
library("knitr")
library("VennDiagram")
library("readr")
```

Before any analysis, we need to specify the ADAGE model and the data compendium
we want to use.

```{r}
model <- eADAGEmodel
compendium <- PAcompendium
probe_dist <- probedistribution
```

Let's load in the sample dataset that come with the package. It's expression data
from *Pseudomonas aeruginosa* wild type and *∆anr* grown as biofilms on ∆F508 cystic
fibrosis bronchial epithelial cells (CFBEs). Detailed information about
the dataset can be found here
[GSE67006](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE67006).
All its CEL files are stored in the folder "./inst/extdata/anr".
To load your own dataset, simply modify this input path.

```{r}
input_path <- system.file("extdata", "anr/", package = "ADAGEpath")
data_raw <- load_dataset(input = input_path, isProcessed = FALSE,
                         isRNAseq = FALSE, model = model,
                         compendium = compendium, quantile_ref = probe_dist,
                         norm01 = FALSE)
```

ADAGE only accepts expression values in the (0,1) range. We linearly transform
expression values to be between 0 and 1 using the Pa compendium as the reference.

```{r}
data_normed <- zeroone_norm(input_data = data_raw, use_ref = TRUE,
                            ref_data = compendium)
```

Now let's specify the phenotypes for each sample. It needs to be a character
vector and has the same sample order as the expression data loaded above.

```{r}
data_pheno <- c("mt", "mt", "mt", "wt", "wt", "wt")
```

# ADAGE signature analysis

## Activity calculation
We calculate the activity of each signature for each sample in the dataset.

```{r}
data_activity <- calculate_activity(input_data = data_normed, model = model)
```

The returned `data_activity` is a `data.frame` with signature names in the first
column and activity values per sample starting from the second column.

## Active signature detection
We want to find signatures that are differentially active between
*anr* mutant and wildtype samples.
We use [limma](https://bioconductor.org/packages/release/bioc/html/limma.html)
to perform a differential activation test. limma is more robust
than t-test when sample size is small. A two-group limma analysis is
provided in the function `build_limma()`. You can also build other limma models
to test signatures' activities when the experimental design is more complex.

In the limma test, we will use "wt" as the control phenotype.
Because there are a lot of signatures passing the significance cutoff, here we
use the more stringent Bonferroni procedure instead of the Benjamini–Hochberg
procedure for multiple hypothesis correction.

```{r}
limma_result <- build_limma(data_activity, phenotypes = data_pheno,
                            control_pheno = "wt",
                            use.bonferroni = TRUE)
```

To take both absolute activity difference and significance into account, we
use pareto fronts to pick the most differentially active signatures. We extract
differentially active signatures in the first 10 layers of pareto
fronts. Modify `N_fronts` to get more or fewer signatures.

```{r}
active_sigs <- get_active_signatures(limma_result = limma_result,
                                     pheno_group = "both", 
                                     method = "pareto", N_fronts = 10)
```

Signatures that are differentially active between *anr* mutant and wildtype are:

```{r}
active_sigs
```

Plot each signature's activity change and significance in the `limma` test.

```{r}
plot_volcano(limma_result, highlight_signatures = active_sigs,
             interactive = TRUE)
```

Look at how the activities of active signature vary across samples.

```{r}
plot_activity_heatmap(activity = data_activity, signatures = active_sigs)
```

Combining the volcano plot and the activity heatmap, Node35pos is the most
active signature in anr mutant, followed by Node233pos, Node140pos.
On the other side, Node38neg, Node31pos, Node269pos, Node205neg are most active
in wildtype.

## Overlapping signature removal
To reduce the number of signatures to look at, we can check whether these active
signatures overlap with each other.

`plot_signature_overlap` creates a heatmap of odds ratios. The odds ratio
represents the odds that two signatures share a specific number of genes.

```{r}
signature_similarity <- plot_signature_overlap(selected_signatures = active_sigs,
                                               model = model)
```

Next we calculate the marginal activities of similar signatures. Marginal
activity is defined as the activity of signature A after removing genes that it
shares with signature B.

```{r}
marginal_activity <- calculate_marginal_activity(input_data = data_normed,
                                                 selected_signatures = active_sigs,
                                                 model = model)
```

Again, we build a limma model to test whether these marginal activities
are still strongly different between two conditions.

```{r}
marginal_limma <- build_limma(input_data = marginal_activity,
                              phenotypes = data_pheno, control_pheno = "wt")
```

Let's visualize the marginal activities in a matrix heatmap.  The value in this
matrix represents the -log10 transformed adjusted p-value in
the activation test when the effect of the column signature is removed from
the row signature. Values in the diagonal of the heatmap are the activation
significance of signatures themselves. Activation significance below the
significance cutoff is marked by a cross sign.

```{r}
plot_marginal_activation(marginal_limma_result = marginal_limma,
                         signature_order = colnames(signature_similarity),
                         sig_cutoff = 0.05)

```

Based on this plot, we can see that Node119pos, Node214neg, and Node299pos are
completely masked by Node191pos, because after removing Node191pos from them,
they become non-significant. Node130pos and Node250neg are masked by
Node31pos. Node285neg is masked by Node205neg.
Node154pos is masked by Node57neg. Node63neg, Node39neg, Node228pos, Node158neg,
Node140pos, Node269neg and Node31neg are masked by Node35pos. Node185neg and
Node275pos are masked by Node67pos. Node278pos is masked by Node9pos.
We can see that Node67pos, Node35pos, and Node233pos each has unique
genes that make them still significant even after removing the effect
of another signature. We can safely remove a signature being masked by another
signature as long as we keep the second signature.

```{r}
unique_active_sigs <- remove_redundant_signatures(marginal_limma,
                                                  sig_cutoff = 0.05)
unique_active_sigs
```

Check out each signature's activity change and significance after removing
overlapping signatures.

```{r}
plot_volcano(limma_result, highlight_signatures = unique_active_sigs,
             interactive = TRUE)
```

Look at how the activities of active signature vary across samples after removing
overlapping signatures.

```{r}
plot_activity_heatmap(activity = data_activity, signatures = unique_active_sigs)
```

## Active signature interpretation
### Genes in signatures
First of all, we can directly inspect gene composition of a signature or a group
of signatures.

```{r}
DT::datatable(annotate_genes_in_signatures(selected_signatures = "Node35pos",
                                           model = model))
```

### Associate pathways
Next we can associate existing pathways to signatures. Here we use KEGG pathways.
We retrive *Pseudomonas aeruginosa* KEGG terms from the
[TRIBE](http://tribe.greenelab.com/#/home) webserver. TRIBE periodically updates
KEGG terms. To reproduce results presented in the paper, we use the KEGG
pathways that were retrieved from TRIBE on Aug. 12 2015. (not implemented yet)
You can also repeat the analysis with up-to-date KEGG terms from TRIBE.
TRIBE also supports Gene Ontology terms. You can simply replace "KEGG" with "GO"
in the following steps to repeat the analysis with GO terms.

```{r}
KEGG <- fetch_geneset(type = "KEGG")
# we only consider KEGG pathways with more than 5 genes and less than 100 genes
# as meaningful pathways
KEGG_subset <- KEGG[lengths(KEGG) >= 5 & lengths(KEGG) <= 100]
```

We associate active signatures to known KEGG pathways.

```{r}
pathway_association <- annotate_signatures_with_genesets(
  selected_signatures = unique_active_sigs, model = model,
  genesets = KEGG_subset)
```

Calculate the activity of associated pathways inside active signatures. This help
differentiate active vs. non-active pathways assoicated with active signatures.

```{r, fig.height= 10, fig.width=10}
pathway_activity <- signature_geneset_activity(
  signature_geneset_df = pathway_association[, c("signature", "geneset")],
  gene_set_list = KEGG_subset, model = model, input_data = data_normed)
plot_activity_heatmap(pathway_activity)
```

We run a limma test on pathway activities and find pathways that are truly
active in this dataset.

```{r}
pathway_limma <- build_limma(pathway_activity, phenotypes = data_pheno,
                             control_pheno = "wt", use.bonferroni = TRUE)
# combine pathway association and pathway activation test results
combined_result <- combine_geneset_outputs(
  signature_geneset_association = pathway_association,
  geneset_limma_result = pathway_limma)
knitr::kable(combined_result, digits = 4, row.names = FALSE, align = "c")
```

### Gene-gene networks
We next visualize how genes in the active signatures cluster in the
ADAGE-derived gene-gene network.

We calculate an expression fold change for each gene and pass it to the
gene-gene network to show as node color. Again, we use limma to test
differential expression and get the logFC.

```{r}
data_raw_limma <- build_limma(input_data = data_raw, phenotypes = data_pheno,
                              control_pheno = "wt")
# build a gene:fold change table from limma result
gene_logFC <- data.frame(geneID = rownames(data_raw_limma),
                         logFC = data_raw_limma$logFC)
```

Visualize the ADAGE gene-gene network of the active signatures.

```{r}
visualize_gene_network(selected_signatures = unique_active_sigs,
                       gene_color_value = gene_logFC,
                       model = model, cor_cutoff = 0.5,
                       curated_pathways = KEGG)
```

We can also visualize one signature at a time. 

```{r}
visualize_gene_network(selected_signatures = "Node35pos",
                       gene_color_value = gene_logFC,
                       model = model, cor_cutoff = 0.5,
                       curated_pathways = KEGG)
```

# A comparison between ADAGEpath and GSEA
We also perform a GSEA analysis on this dataset.

```{r}
# specifiy file path to the GSEA java program
GSEA_file <- "gsea2-2.2.4.jar"
# prefix of the folder that stores GSEA results
GSEA_output_name <- "GSEA_result"
# folder to store GSEA data and results
GSEA_folder <- "GSEA_data"

# NOTICE!!! The folder cannot already exist.
if (dir.exists(GSEA_folder)) {
  stop("The GSEA folder already exists. Please give a new folder name.")
}
dir.create(GSEA_folder, showWarnings = TRUE)


if (file.exists(GSEA_file)){

  # create the class lable file for gsea
  label_file <- file.path(GSEA_folder, "label.cls")
  write(paste(length(data_pheno), length(unique(data_pheno)), 1, sep = " "),
        file = label_file)
  write("# con1 con2", file = label_file, append = TRUE)
  write(paste(data_pheno, collapse = " "), label_file, append=TRUE)
  
  # create the gene expression file for gsea
  expression_file <- file.path(GSEA_folder, "expression.txt")
  readr::write_tsv(data_raw, expression_file, col_names = TRUE)

  # create the gene set file for gsea
  geneset_file <- file.path(GSEA_folder, "geneset.gmt")
  write(paste(names(KEGG_subset)[1], "na", paste(unlist(KEGG_subset[[1]]),
                                                 collapse = "\t"), sep = "\t"),
        file = geneset_file)
  invisible(sapply(2:length(KEGG_subset),
                   function(x) write(paste(names(KEGG_subset)[x], "na",
                                           paste(unlist(KEGG_subset[[x]]),
                                                 collapse = "\t"), sep = "\t"),
                                     file = geneset_file, append = TRUE)))

  # run GSEA
  gsea <- paste0("java -Xmx512m -cp ", GSEA_file, " xtools.gsea.Gsea -res ",
                expression_file," -cls ",label_file," -gmx ",geneset_file,
                " -collapse false -mode Max_probe -norm meandiv -nperm 1000 ",
                "-permute gene_set -rnd_type no_balance -scoring_scheme weighted ",
                "-rpt_label ", GSEA_output_name ," -metric Signal2Noise -sort real ",
                "-order descending -include_only_symbols true -make_sets true ",
                "-median false -num 100 -plot_top_x 20 -rnd_seed timestamp ",
                "-save_rnd_lists false -set_max 100 -set_min 5 -zip_report false ",
                "-out ", GSEA_folder, " -gui false")
  system(gsea)

  # read in GSEA results
  gsea_con1 <- readr::read_tsv(Sys.glob(
    file.path(GSEA_folder, paste0(GSEA_output_name, ".Gsea*"),
              "gsea_report_for_con1_*.xls")), col_names = TRUE)
  gsea_con2 <- readr::read_tsv(Sys.glob(
    file.path(GSEA_folder, paste0(GSEA_output_name, ".Gsea*"),
              "gsea_report_for_con2_*.xls")), col_names = TRUE)

  # get significant results
  gsea_con1_sig <- gsea_con1[gsea_con1$`FDR q-val` <= 0.05, ]
  gsea_con2_sig <- gsea_con2[gsea_con2$`FDR q-val` <= 0.05, ]
  KEGG_GSEA <- unique(c(gsea_con1_sig$NAME, gsea_con2_sig$NAME))

} else {
  stop("Please make sure the file path to the GSEA java program is correct!")

}

```

Now let's compare the significant KEGG pathways in GSEA with the KEGG pathways
associated with the active signatures.

```{r}
# GSEA automatically converts pathway names to upper case, so we also convert
# the pathways identified by ADAGEpath.
KEGG_ADAGE <- toupper(as.character(unique(pathway_association$geneset)))

# make a venn diagram for pathway overlap
grid.newpage()
invisible(draw.pairwise.venn(length(KEGG_GSEA), length(KEGG_ADAGE),
                             length(intersect(KEGG_GSEA, KEGG_ADAGE)),
                             c("GSEA", "ADAGE"), fill = c("red", "blue")))
```

As we can see, GSEA and ADAGEpath agree on many pathways. Here we look into
pathways significantly enriched in GSEA but not associated with chosen active
signatures. We check whether they are assoicated with any ADAGE signatures and
how differentially active the associated signatures are.

```{r}
# built a match table between normal case and upper case
match_table <- data.frame(normal_case = names(KEGG_subset),
                          upper_case = toupper(names(KEGG_subset)),
                          stringsAsFactors = FALSE)

# pathways enriched in GSEA but not associated with ADAGE signatures
# in the top 10 layers of pareto fronts
GSEA_only <- match_table$normal_case[match(setdiff(KEGG_GSEA, KEGG_ADAGE),
                                           match_table$upper_case)]

# check whether GSEA-only KEGG pathways are associated with any signatures
# this step will take a while because it checks association between all signatures
# and all pathways
associated_sigs <- find_associated_signatures(input_genesets = GSEA_only,
                                              model = model,
                                              gene_set_list = KEGG_subset,
                                              significance_cutoff = 0.05,
                                              signature_limma_result = limma_result)
associated_sigs <- unlist(associated_sigs)

# see how these signatures perform in the activation test
plot_volcano(limma_result, highlight_signatures = associated_sigs,
             interactive = TRUE)
```

Here are pathways enriched in GSEA but not associated with any signatures:

```{r}
GSEA_only[is.na(associated_sigs)]
```

Here are pathways that were detected to be active in ADAGEpath but do not pass
the significance cutoff in GSEA. Let's see how they perform in GSEA.

```{r}
ADAGE_active <- unique(combined_result$geneset[combined_result$adj.P.Val <= 0.05])
ADAGE_only <- setdiff(toupper(ADAGE_active), KEGG_GSEA)
ADAGE_only_GSEA <- rbind(gsea_con1[na.omit(match(ADAGE_only, gsea_con1$NAME)), ],
                         gsea_con2[na.omit(match(ADAGE_only, gsea_con2$NAME)), ])
knitr::kable(ADAGE_only_GSEA)
```

There are also signatures uncharacterized by KEGG (not associated with any
exisiting KEGG pathway).

```{r}
uncharacterized_sigs <- setdiff(unique_active_sigs, pathway_association$signature)
uncharacterized_sigs
```

Check the signature similarity of the uncharacterized signatures.

```{r}
plot_signature_overlap(selected_signatures = uncharacterized_sigs, model = model)
```

Visualize uncharacterized signatures in the gene-gene network.

```{r}
visualize_gene_network(selected_signatures = uncharacterized_sigs,
                       gene_color_value = gene_logFC,
                       model = model, cor_cutoff = 0.5,
                       curated_pathways = KEGG)
```